\ttitle

### 2014-01-14 Chapters/Chapter3/one_class_svm.tex
Feedback Chapter 4

% !TEX root = ../../main.tex
\section{One-Class Support Vector Machine}\label{sec:one_class_svm}
In this section we will discuss the details of a \gls{svm} and \gls{oc-svm} implementations.
The classical implementation of a \gls{svm} is to classify a dataset in two distinct classes.
This is a common use case, although sometimes there is no training data for both classes available.
Still, one would like to classify new data points as regular, in-class, or out-of-class, \eg in the case of a novelty detection.
With that problem only data examples from one class are available and the objective is to recognize new data points that are not part of that class.
This unsupervised learning problem is closely related to density estimation.
In that context, the problem can be the following.
Assume an underlying probability distribution $P$ and a data point $\vectorsym{x}$ drawn from this distribution.
The goal is to find a subset $S$ of the input space, such that the probability that $\vectorsym{x}$ lies inside of $S$ equals some predetermined value $\nu$ between $0$ and $1$ \cite{scholkopf1999support}.

@@

A \gls{svm} can create a boundary between linear-separable data points, making it a binary linear classifier.
More flexible non-linear boundaries can be obtained by the use of a non-linear function $\phi(x)$, as illustrated in \Cref{fig:kernel_mapping}.
This function maps the input data from space $\mathcal{I}$ to a higher dimensional space $\mathcal{F}$.
The \gls{svm} can create a linear separating hyperplane in the space $\mathcal{F}$ that separates the data points from the two classes.
When the hyperplane is projected on to the (lower) original input space $\mathcal{I}$ it creates a non-linear separating curve.
A schematic overview of this process is illustrated in \Cref{fig:svm_mapping_spaces}.
The original data points $x \in \mathcal{I}$ are projected by $\phi(x)$ to the points $z \in \mathcal{F}$.
In that space the linear hyperplane is constructed by weights $w$.
This results in a non-linear hyperplane $y$ in input space $\mathcal{I}$.
The mapping and projection of data points can be efficient (and implicit) performed by using the kernel trick, which is discussed in \Cref{subsec:kernels}.

\begin{figure}
\centering
  \includegraphics[width=0.8\textwidth]{./Figures/chapter3/svm_kernel_mapping.png}
  \caption[Kernel mapping]{Visualization of a \gls{svm} in 2D.
  The non-linear boundary in the input space $\mathcal{I}$ (left) is transformed to a linear boundary in the feature space $\mathcal{F}$ (right) by mapping the data points with the function $\phi$.
  The kernel trick uses a function $K$ which performs an implicit mapping. Image from Wikipedia.org}
  \label{fig:kernel_mapping}
\end{figure}

@@

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{./Figures/chapter3/svm_separating_plane_with_margin.png}
  \caption[\gls{svm} and the separating hyperplane]{Illustration of the separating hyperplane of a \gls{svm}.
  Here $w$ is the normal vector for the separating hyperplane and size of the margin is $\frac{2}{\lVert{w}\rVert}$.
  Image from Wikipedia.org}
  \label{fig:svm_hyperplane}
\end{figure}

With every classification method there is a risk of overfitting.
In that case the random error or noise of the data set is described instead of the underlying data.
The \gls{svm} classifier can use a \emph{soft margin} by allowing some data points to lie within the margin, instead of on the margin or farther away from the hyperplane.
For this it introduces \emph{slack variables} $\xi_i$ for each data point and the constant $C > 0$ determines the trade-off between maximizing the margin and the number of data points within that margin (and thus the training errors).
The slack variables are a generalization to minimize the \emph{sum of deviations}, rather than the \emph{number} of incorrect data points \cite{cherkassky2007learning}\footnote{if $\xi_i$ is chosen to be an indicator function, it still minimizes the \emph{number} of incorrect data points.}.
The objective function for a \gls{svm} is the following minimization function:

@@

The number of basis functions, the center parameters that correspond with support vectors, and the weights in the output layer are all automatically determined via the optimal hyperplane in features space $\mathcal{F}$ \cite{cherkassky2007learning}.
The width parameter $\sigma$ is equal for all basis functions and is set a priori and determines the flexibility and complexity of the boundary.
In \Cref{subsec:svm_model_parameters} this (hyper)parameter for a \gls{svm} is further discussed.

@@

In the maximization of \Cref{eq:svdd_lagrange} a large fraction of the multipliers $\alpha_i$ become zero and for a small fraction $\alpha_i > 0$.
This small fraction, for which $\alpha_i$ is non-zero, are called the \glspl{sv} and these objects lie on the boundary of the description.
The center of the hypersphere only depends on this small number of \glspl{sv} and the objects for which $\alpha_i = 0$ can be disregarded from the solution.
Testing the membership of a (new) object $\vectorsym{z}$ is done by determining if the distance to the center $\vectorsym{a}$ of the sphere is equal or smaller to the radius $R$:
\begin{equation}\label{eq:svdd_test_object}
  \lVert \vectorsym{z} - \vectorsym{a} \rVert ^2 = (\vectorsym{z} \cdot \vectorsym{z}) - 2 \sum_{i} \alpha_i(\vectorsym{z} \cdot \vectorsym{x}_i) + \sum_{i,j}(\vectorsym{x}_i \cdot \vectorsym{x}_j) \le R^2
\end{equation}
As with \Cref{eq:svm_lagrange}, the solution of this equation only relies on dot products between the data points in $\vectorsym{x}$ and $\vectorsym{z}$.
This means that the kernel projection and trick, as discussed in \Cref{subsec:kernels}, can be applied to \gls{svdd} as well \cite{tax1999support,tax2002uniform}.

>>>>

% !TEX root = ../../main.tex
\section{One-Class Support Vector Machine}\label{sec:one_class_svm}
In this section we will discuss the details of an \gls{svm} and \gls{oc-svm} implementations.
The classical implementation of an \gls{svm} is to classify a dataset in two distinct classes.
This is a common use case, although sometimes there is no training data for both classes available.
Still, one would like to classify new data points as regular, in-class, or out-of-class, \eg in the case of a novelty detection.
With that problem only data examples from one class are available and the objective is to recognize new data points that are not part of that class.
This unsupervised learning problem is closely related to density estimation.
In that context, the problem can be the following.
Assume an underlying probability distribution $P$ and a data point $\vectorsym{x}$ drawn from this distribution.
The goal is to find a subset $S$ of the input space, such that the probability that $\vectorsym{x}$ lies inside of $S$ equals some predetermined value $\nu$ between $0$ and $1$ \cite{scholkopf1999support}.

@@

An \gls{svm} can create a boundary between linear-separable data points, making it a binary linear classifier.
More flexible non-linear boundaries can be obtained by the use of a non-linear function $\phi(x)$, as illustrated in \Cref{fig:kernel_mapping}.
This function maps the input data from space $\mathcal{I}$ to a higher dimensional space $\mathcal{F}$.
The \gls{svm} can create a linear separating hyperplane in the space $\mathcal{F}$ that separates the data points from the two classes.
When the hyperplane is projected on to the (lower) original input space $\mathcal{I}$ it creates a non-linear separating curve.
A schematic overview of this process is illustrated in \Cref{fig:svm_mapping_spaces}.
The original data points $x \in \mathcal{I}$ are projected by $\phi(x)$ to the points $z \in \mathcal{F}$.
In that space the linear hyperplane is constructed by weights $w$.
This results in a non-linear hyperplane $y$ in input space $\mathcal{I}$.
The mapping and projection of data points can be efficient (and implicit) performed by using the kernel trick, which is discussed in \Cref{subsec:kernels}.

\begin{figure}
\centering
  \includegraphics[width=0.8\textwidth]{./Figures/chapter3/svm_kernel_mapping.png}
  \caption[Kernel mapping]{Visualization of an \gls{svm} in 2D.
  The non-linear boundary in the input space $\mathcal{I}$ (left) is transformed to a linear boundary in the feature space $\mathcal{F}$ (right) by mapping the data points with the function $\phi$.
  The kernel trick uses a function $K$ which performs an implicit mapping. Image from Wikipedia.org}
  \label{fig:kernel_mapping}
\end{figure}

@@

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{./Figures/chapter3/svm_separating_plane_with_margin.png}
  \caption[\gls{svm} and the separating hyperplane]{Illustration of the separating hyperplane of an \gls{svm}.
  Here $w$ is the normal vector for the separating hyperplane and size of the margin is $\frac{2}{\lVert{w}\rVert}$.
  Image from Wikipedia.org}
  \label{fig:svm_hyperplane}
\end{figure}

With every classification method there is a risk of overfitting.
In that case the random error or noise of the data set is described instead of the underlying data.
The \gls{svm} classifier can use a \emph{soft margin} by allowing some data points to lie within the margin, instead of on the margin or farther away from the hyperplane.
For this it introduces \emph{slack variables} $\xi_i$ for each data point and the constant $C > 0$ determines the trade-off between maximizing the margin and the number of data points within that margin (and thus the training errors).
The slack variables are a generalization to minimize the \emph{sum of deviations}, rather than the \emph{number} of incorrect data points \cite{cherkassky2007learning}\footnote{if $\xi_i$ is chosen to be an indicator function, it still minimizes the \emph{number} of incorrect data points.}.
The objective function for an \gls{svm} is the following minimization function:

@@

The number of basis functions, the center parameters that correspond with support vectors, and the weights in the output layer are all automatically determined via the optimal hyperplane in features space $\mathcal{F}$ \cite{cherkassky2007learning}.
The width parameter $\sigma$ is equal for all basis functions and is set a priori and determines the flexibility and complexity of the boundary.
In \Cref{subsec:svm_model_parameters} this (hyper)parameter for an \gls{svm} is further discussed.

@@

In the maximization of \Cref{eq:svdd_lagrange} a large fraction of the multipliers $\alpha_i$ become zero and for a small fraction $\alpha_i > 0$.
This small fraction, for which $\alpha_i$ is non-zero, are called the \glspl{sv} and these objects lie on the boundary of the description.
The center of the hypersphere only depends on this small number of \glspl{sv} and the objects for which $\alpha_i = 0$ can be discarded from the solution.
Testing the membership of a (new) object $\vectorsym{z}$ is done by determining if the distance to the center $\vectorsym{a}$ of the sphere is equal or smaller to the radius $R$:
\begin{equation}\label{eq:svdd_test_object}
  \lVert \vectorsym{z} - \vectorsym{a} \rVert ^2 = (\vectorsym{z} \cdot \vectorsym{z}) - 2 \sum_{i} \alpha_i(\vectorsym{z} \cdot \vectorsym{x}_i) + \sum_{i,j}(\vectorsym{x}_i \cdot \vectorsym{x}_j) \le R^2
\end{equation}
As with \Cref{eq:svm_lagrange}, the solution of this equation only relies on dot products between the data points in $\vectorsym{x}$ and $\vectorsym{z}$.
This means that the kernel projection and trick, as discussed in \Cref{subsec:kernels}, can be applied to \gls{svdd} as well \cite{tax1999support,tax2002uniform}.

### 2014-01-14 Chapters/Chapter4/change_detection.tex
Put captions above tables
Feedback Chapter 4

% !TEX root = ../../main.tex
\section{Change Detection}\label{sec:method_change_detection}
In the previous sections we have discussed what data is used for the model construction, and how that model is interpreted in the context of change detection.
We have shown how the multi-dimensional signal is reduced to a single dimension time series data, based on the extracted radius $R$ from the constructed hypersphere.
This metric is used to discover changes in the underlying data generation process.
We argued that in the case of the data objects set becomes more heterogeneous, we expect an increase in the radius $R$.
In the same manner we expect that when the change is in the oldest data objects of the working set the radius $R$ will decrease, since the data becomes more homogeneous.
An abstract representation of these expectations is illustrated in \Cref{fig:radius_expectation}.

In this section we discuss the methods applicable to the extracted radius $R$ of the hypersphere, in order to discover a change in the underlying distribution properties.
This employs the second stage of our algorithm, analogue to the second stage of the unifying framework by Takeuchi and Yamanishi~\cite{takeuchi2006unifying}.
It is also based on the \gls{svcpd} method by Camci~\cite{camci2010change}, although we have slightly simplified the algorithm.

In the \gls{svcpd} the first step for a new data point $\vectorsym{z}$ is to check whether the data point lies within the enclosing (hyperspherical) boundary of the constructed model.
This will label the new data point as an in- or outlier, realtive to the current data set.
Depending on that outcome, the model is updated to include the new data point.
The second step is to inspect the (changed) model properties, \ie the radius $R$.
In our method we only employ the second step: the inspection of the radius $R$.
We regard the first step (determine if $\vectorsym{z}$ is an outlier) to be superfluent: in case it is an outlier, the model update step will change the radius $R$ in order to let the boudary include the new data point $\vectorsym{z}$.
Would the new data object $\vectorsym{z}$ be already an inlier, then the updated model would not have a significant change in radius $R$.
Our algorithm is outlined in \Cref{tab:algorithm_proposed_method}.

@@

\begin{center}\begin{table}
  \begin{tabular}{ l p{12cm} }
    \hline
    Step & Action \\
    \hline
    1 & Start with $n$ data objects as working set and construct hypersphere \\
    2 & Add next data object $x_t$ and drop first one \\
    3 & Identify new hyper-sphere and its \emph{approximate radius} \\
    4 & calculate radius average of hyperspheres since last change point \\
    5 & calculate radius ratio $\hbar$. If lower than ${th}_{low}$ or greater than ${th}_{high}$ then mark $t$ as change point \\
    6 & continue from step 2, until no more data objects available \\
    7 & Apply post-processing: merge close change points \\
    \hline
  \end{tabular}
  \caption[Proposed algorithm]{Algorithm of the proposed change detection method, using the Ratio-Thresholding method of \ref{subsec:ratio_thresholding} for change detection.}
  \label{tab:algorithm_proposed_method}
\end{table}\end{center}

\subsection{CUSUM based methods}
Many of the simple change detection methods are based on \gls{cusum}, originally introduced by Page~\cite{page1954continuous}.
Other variations and extensions of the method are proposed and used (\cite{inclan1994use,alippi2006adaptive,hsu2007mosum}).
Here we will discuss the simple form, which only detects changes as an increase in value of the examples, in this case the radius $R$.
The cumulative sum for all the values is calculated:
\begin{equation}
  S_n = \sum_{k=1}^n r_k
\end{equation}
and a change is detected when the value of the cumulative sums minus the minimum encountered exceeds a predetermined threshold $h$:
\begin{equation}
  S_n - \operatorname*{min}_{0 \le i < n} S_i \ge h.
\end{equation}

Extensions of the \gls{cusum} method have been proposed.
Amongst others, extensions which extend its use to two-threshold methods, and methods appropiate for changes in mean or variance \cite{inclan1994use}.
Whilst adaptive methods have been proposed \cite{alippi2006adaptive}, most methods require manual determination of the threshold parameter $h$.

@@

The appropriate values for the thresholds are highly depended on the characteristics of the data and the trade-off parameter $C$ (as discussed in \Cref{subsec:oc-svm-svdd}).
This trade-off parameters regulates the fraction of data objects that will be rejected from the constructed model.
Since a high value of $C$ assigns a high penalty to outliers, data objects are more likely to be incorporated into the hypersphere by increasing the radius.
For low values of $C$ the cost of rejecting data objects is, compared to the benefits of a smaller hypersphere, relatively low.
This gives a relation between the parameters $C$, $th_\text{low}$, $th_\text{high}$, and the sensitivity of the radius ratio based thresholding method.
A high value of $C$, or values close to $1$ for $th_\text{low}$ and $th_\text{high}$ result in a sensitive change detection procedure.
The reverse results in less sensitive methods.
The correct values for the intended sensitivity need to be empirically determined.

@@

\subsection{Post-processing}
In our method we use the ratio based thresholding, since it is also used by Camci~\cite{camci2010change}.
From preliminary experiments we observed that a single change in the data can cause many detected change points, using the ratio based method.
This phenomenon can be explained by the nature of the thresholding method.
Since a single (high or low) threshold is set for ratio of the radii, high values are considered to be a change point before the highest point of the peak.
As illustrated in \Cref{fig:tresholding}, the peak at $t_1$ represents a single change point.
The region from $t_2$ to $t_3$ is strictly increasing.
At $t_2$ a change point is detected, since $\hbar_t$ exceeds the threshold.
But since it is increasing up to $t_3$, all values of $\hbar_t$ in that period will trigger the ratio based method to indicate a change point.

>>>>

% !TEX root = ../../main.tex
\section{Change Detection}\label{sec:method_change_detection}
In the previous sections we have discussed what data is used for the model construction, and how that model is interpreted in the context of change detection.
We have shown how the multi-dimensional signal is reduced to a single dimension time series data, based on the extracted radius $R$ from the constructed hypersphere.
This metric is used to discover changes in the underlying data generation process.
We argued that in the case the data objects set becomes more heterogeneous, we expect an increase in the radius $R$.
In the same manner we expect that when the change is in the oldest data objects of the working set, the radius $R$ will decrease, since the data becomes more homogeneous.
An abstract representation of these expectations is illustrated in \Cref{fig:radius_expectation}.

In this section we discuss the methods applicable to the extracted radius $R$ of the hypersphere, in order to discover a change in the underlying distribution properties.
It is the second stage of our algorithm, analogue to the second stage of the unifying framework by Takeuchi and Yamanishi~\cite{takeuchi2006unifying}.
It is also based on the \gls{svcpd} method by Camci~\cite{camci2010change}, although we have simplified the algorithm.

\begin{table}
\begin{center}
  \caption[Proposed algorithm]{Algorithm of the proposed change detection method, using the Ratio-Thresholding method of \ref{subsec:ratio_thresholding} for change detection.}
  \begin{tabular}{ l p{12cm} }
    \hline
    Step & Action \\
    \hline
    & \textbf{Input:} time series $\vectorsym{x}$, window size $n$ \\
    & \textbf{Output:} collection of change points $\vectorsym{t}$ \\
    1 & Start with $n$ data objects as working set and construct hypersphere \\
    2 & Add next data object $x_t$ and drop first one \\
    3 & Identify new hyper-sphere and its \emph{approximate radius} \\
    4 & Calculate radius average of hyperspheres since last change point \\
    5 & Calculate radius ratio $\hbar$. \newline
    If $\hbar$ is  lower than ${th}_{low}$ or greater than ${th}_{high}$ then mark $t$ as change point in $\vectorsym{t}$ \\
    6 & Continue from step 2, until no more data objects available \\
    7 & Apply post-processing: merge close change points \\
    \hline
  \end{tabular}
  \label{tab:algorithm_proposed_method}
\end{center}
\end{table}

\TODO{Algoritme meer in pseudo-code zetten. Prominenter maken.}

In the \gls{svcpd} the first step for a new data point $\vectorsym{z}$ is to check whether the data point lies within the enclosing (hyperspherical) boundary of the constructed model.
This will label the new data point as an in- or outlier, relative to the current data set.
Depending on that outcome, the model is updated to include the new data point.
The second step is to inspect the (changed) model properties, \ie the radius $R$.
In our method we only employ the second step: the inspection of the radius $R$.
We regard the first step (determine if $\vectorsym{z}$ is an outlier) to be superfluent: in case it is an outlier, the model update step will change the radius $R$ in order to let the boundary include the new data point $\vectorsym{z}$.
Would the new data object $\vectorsym{z}$ be already an inlier, then the updated model would not have a significant change in radius $R$.
Our algorithm is outlined in \Cref{tab:algorithm_proposed_method}.

@@

\subsection{CUSUM based methods}
Many of the simple change detection methods are based on \gls{cusum}, originally introduced by Page~\cite{page1954continuous}.
Other variations and extensions of the method are proposed and used (\cite{inclan1994use,alippi2006adaptive,hsu2007mosum}).
Here we will discuss the simple form, which only detects changes as an increase in value of the examples, in this case the radius $R$.
The cumulative sum for all the values is calculated:
\begin{equation}
  S_n = \sum_{k=0}^n R_k
\end{equation}
and a change is detected when the value of the cumulative sums minus the minimum encountered exceeds a predetermined threshold $h$:
\begin{equation}
  \left\{ S_n - \operatorname*{min}_{0 \le i < n} S_i \right\} \ge h.
\end{equation}

Extensions of the \gls{cusum} method have been proposed.
Amongst others, extensions which extend its use to two-threshold methods, and methods appropriate for changes in mean or variance \cite{inclan1994use}.
Whilst adaptive methods have been proposed \cite{alippi2006adaptive}, most methods require manual determination of the threshold parameter $h$.

@@

The appropriate values for the thresholds strongly depend on the characteristics of the data and the parameter $C$ (as discussed in \Cref{subsec:oc-svm-svdd}).
This trade-off parameter regulates the fraction of data objects that will be rejected from the constructed model.
Since a high value of $C$ assigns a high penalty to outliers, data objects are more likely to be incorporated into the hypersphere by increasing the radius.
For low values of $C$ the cost of rejecting data objects is, compared to the benefits of a smaller hypersphere, relatively low.
This gives a relation between the parameters $C$, $th_\text{low}$, $th_\text{high}$, and the sensitivity of the radius ratio based thresholding method.
A high value of $C$, or values close to $1$ for $th_\text{low}$ and $th_\text{high}$ result in a sensitive change detection procedure.
The reverse results in less sensitive methods.
The correct values for the intended sensitivity need to be empirically determined.

@@

\subsection{Post-processing}
In our method we use the ratio based thresholding, also used by Camci~\cite{camci2010change}.
From preliminary experiments we observed that a single change in the data can cause many detected change points, using the ratio based method.
This phenomenon can be explained by the nature of the thresholding method.
Since a single (high or low) threshold is set for ratio of the radii, high values are considered to be a change point before the highest point of the peak.
As illustrated in \Cref{fig:tresholding}, the peak at $t_1$ represents a single change point.
The region from $t_2$ to $t_3$ is strictly increasing.
At $t_2$ a change point is detected, since $\hbar_t$ exceeds the threshold.
But since it is increasing up to $t_3$, all values of $\hbar_t$ in that period will trigger the ratio based method to indicate a change point.

### 2014-01-14 Chapters/Chapter4/data_gathering.tex
Put captions above tables
Feedback Chapter 4

% ------
\subsection{Artificial data}\label{subsec:data_gathering_artificial}
In order to provide an objective comparison to other methods, we will use artificial data sets which are also used in the researches on which our method is based.
These are the data sets used by Takeuchi and Yamanishi \cite{takeuchi2006unifying} and Camci \cite{camci2010change}.
Both construct a collection of one-dimensional time series data according to a second order \gls{ar} model:
\begin{equation}
  x_t = a_1 x_{t-1} + a_2 x_{t-2} + \epsilon_t.
\end{equation}
Over the different data series the mean and variance of the Gaussian random variable $\epsilon_t$ differs and changes at pre-determined change points.
Using this data set an objective quality measure over the change detection methods can be obtained and compared.
All the used data sets are listed and analyzed in \Cref{Chapter5}.

% ------
\subsection{Real-world data}\label{subsec:data_gathering_real_world}
In the second type of data sets we apply our method of change detection and temporal segmentation to real-world data sets.
For our setup we record the activities of humans performed both in- and outdoor in an uncontrolled environment.
Activities performed include sitting, standing, walking, running in a straight and curved line, and walking up- and downstairs.
Our method uses the signals from the accelerometer, magnetic field, and rotation sensors.
These time series data are used to detect change points.
A video recording from the performed activity is used to annotate the time series with real change points.
The discovered change point are compared with these annotated change points to give a subjective quality measure.
In \Cref{Chapter6} we give a detailed analysis of the performed activities and the recorded data sets.

For the experiments we used a HTC Sensation XE smartphone as recording device.
The activities were recorded using a free Android application \cite{sensorlogger}.
This application was chosen for its convenient data format of the sensor recording and its regularity of the sampling interval.
Table~\ref{tab:recorded_metrics} lists all the recorded metrics.
For our expiments we used the data for the accelerometer, magnatief field and rotation.

\begin{center}\begin{table}
  \begin{tabulary}{\textwidth}{|l|L|c|c|}
    \hline
    Metric & Description & Units of measure & Typical range \\
    \hline \hline
    Accelerometer & Acceleration force along each axis (including gravity). & $m/s^2$ & $-20$ -- $20$ \\
    \hline
    Gravity & Force of gravity along each axis. & $m/s^2$ & $-10$ -- $10$\\
    \hline
    Gyroscope & Rate of rotationg around each axis. & $rad/s$ & $-15$ -- $15$\\
    \hline
    Light & Light sensitive sensor at the front of the phone. & & $0$ -- $10000$ \\
    \hline
    Linear acceleration & Acceleration force along each axis (excluding gravity). & $m/s^2$ & $-20$ -- $20$ \\
    \hline
    Magnetic field & Geomagnetic field strength along each axis. & $\mu T$ & $-60$ -- $60$ \\
    \hline
    Orientation & Degrees of rotation around the three physical axis. & Degrees & $-100$ -- $360$ \\
    \hline
    Rotation & Measre of rotation around the device's rotation axis. & Unitless & $-1$ -- $1$\\
    \hline
  \end{tabulary}
  \caption[Measured metrics]{Measured metrics. The set of axis is always the triple (x, y, z) direction.}
  \label{tab:recorded_metrics}
\end{table}\end{center}

>>>>

% ------
\subsection{Artificial data}\label{subsec:data_gathering_artificial}
In order to provide an objective comparison to other methods, we will use artificial data sets which are also used in the earlier work on which our method is based.
These are the data sets used by Takeuchi and Yamanishi \cite{takeuchi2006unifying} and Camci \cite{camci2010change}.
Both construct a collection of one-dimensional time series data according to a second order \gls{ar} model:
\begin{equation}
  x_t = a_1 x_{t-1} + a_2 x_{t-2} + \epsilon_t.
\end{equation}
Over the different data series the mean and variance of the Gaussian random variable $\epsilon_t$ differs and changes at pre-determined change points.
Using this data set an objective quality measure over the change detection methods can be obtained and compared.
All the used data sets are listed and analyzed in \Cref{Chapter5}.

% ------
\subsection{Real-world data}\label{subsec:data_gathering_real_world}
In the second type of data sets we apply our method of change detection and temporal segmentation to real-world data sets.
For our setup we record the activities of humans performed both in- and outdoor in an unknown environment.
Activities performed include sitting, standing, walking, running in a straight and curved line, and walking up- and downstairs.
Our method uses the signals from the accelerometer, magnetic field, and rotation sensors.
These time series data are used to detect change points.
A video recording from the performed activity is used to annotate the time series with real change points.
The discovered change points are compared with these annotated change points to give a subjective quality measure.
In \Cref{Chapter6} we give a detailed analysis of the performed activities and the recorded data sets.

For the experiments we used a smartphone with inertial sensors as recording device.
The activities were recorded using a free \textsc{Android} application \cite{sensorlogger}.
This application was chosen for its convenient data format of the sensor recording and its regularity of the sampling interval.
Table~\ref{tab:recorded_metrics} lists all the recorded metrics.
For our experiments we used the data for the accelerometer, magnetic field and rotation.

We implemented our algorithm in \textsc{MATLAB}.
For the \gls{svdd} model construction phase we used the \textsc{dd\_tools} library by Tax \cite{Ddtools2013}, which depends on the \textsc{PR\_tools} package, originating from the book by Van Der Heijden \etal \cite{van2005classification}.
Alternatively, the widely used \textsc{LibSVM} \cite{chang2011libsvm} add-on for the \gls{svdd} algorithm by Chang \etal \cite{changrevisit} can be used.

\begin{center}\begin{table}
  \caption[Measured metrics]{Measured sensor metrics. The set of axis is always the triple (x, y, z) direction.}
  \begin{tabulary}{\textwidth}{|l|L|c|c|}
    \hline
    Sensor metric & Description & Units of measure & Typical range \\
    \hline \hline
    Accelerometer & Acceleration force along each axis (including gravity). & $m/s^2$ & $-20$ -- $20$ \\
    \hline
    Gravity & Force of gravity along each axis. & $m/s^2$ & $-10$ -- $10$\\
    \hline
    Gyroscope & Rate of rotation around each axis. & $rad/s$ & $-15$ -- $15$\\
    \hline
    Light & Light sensitive sensor at the front of the phone. & & $0$ -- $10000$ \\
    \hline
    Linear acceleration & Acceleration force along each axis (excluding gravity). & $m/s^2$ & $-20$ -- $20$ \\
    \hline
    Magnetic field & Geomagnetic field strength along each axis. & $\mu T$ & $-60$ -- $60$ \\
    \hline
    Orientation & Degrees of rotation around the three physical axis. & Degrees & $-100$ -- $360$ \\
    \hline
    Rotation & Measure of rotation around the device's rotation axis. & Unitless & $-1$ -- $1$\\
    \hline
  \end{tabulary}

  \label{tab:recorded_metrics}
\end{table}\end{center}

### 2014-01-14 Chapters/Chapter4/method_overview.tex
Finalizing Chapter 2, literatur review...
Feedback Chapter 4

% !TEX root = ../../main.tex
\section{Algorithm overview}\label{sec:method_overview}
The proposed method of this thesis follows the unifying framework as introduced by Takeuchi and Yamanishi \cite{takeuchi2006unifying} and an similar implementation by Camci \cite{camci2010change} with \glspl{svm}.
The unifying framework relates the detection of outliers with change points and divides the whole process in two stages.
The first stage determines the outliers in a time series by giving a score based on the deviation from a learned model, and thereby creates a new time series.
The second stage runs on that new created time series and calculates a average over a window of the outlier scores.
The problem of change detection is then reduced to outlier detection over that average-scored time series.
This method is named \gls{changeFinder} by the authors.
The implementation by Camci, which uses \glspl{svm} to detect changes is named \acrlong{svcpd}.

Whereas \gls{changeFinder} uses double probability estimation algorithm, our approach follows \gls{svcpd} by constructing a \gls{svm} over a sliding window.
The \gls{svcpd} algorithm uses the location of new data points in the feature space $\mathcal{F}$ with respect to the hypersphere and the hypersphere's radius $R$ to determine whether the new data point represents a change point.
Our approach is a slight simplification of the algorithm: we only use the updated model properties (\ie the radius $R$) to detect a change.
This difference is further discussed and justified in \Cref{sec:method_change_detection}.

This section gives a description of the method used for the experiments and change detection mechanism.
First described is the method to process the gathered sensor data.
A schematic overview is given in \Cref{fig:method_overview} and shows the steps of the method.
A more detailed explanation of the ``Update model'' step follows.

As graphically represented in \Cref{fig:method_overview}, the change detection method starts by processing the data from sensor, such as the accelerometer, magnetic orientation and rotation metrics.

The first step is to process the raw streams of data originating from a multiple of sensors.
The two processes applied are alignment and normalization.
Due to noisy sampling, not all the timestamps in the data streams are sensed at the same timestamp.
Since the \gls{svdd} method requires all the data stream at every timestamp and can not handle missing data on one of the timestamps, all the unique timestamps are filtered out.
Whilst this results in an overall filtering effect, in practice between $1\%$ and $5\%$ of each data stream is disregarded.
The effect of this filtering is not significant and the data is not modified.

Due to the nature of the sensor signals, a normalization step is required in order to set the weight for all the data streams equal.
The range of the accelerometer signal typically spans $-20$ to $20$, the magnetic field from $-60$ to $60$ and the rotations range is from $-1$ to $1$.
This means that a relative small change in the accelerometer stream could have a much larger impact on the model than the same (absolute) change in the rotation stream, whilst the latter has a larger relative impact.
The normalization step ensures that all data is weighted equally and changes in the data are all proportional.

@@

This final step of this method, step $5$, is the interpretation of the model properties.
Many algorithms can be used for this process, all which take a one-dimensional time series as input and determine where change has occured.
In our setup we used the \gls{rt} and \gls{cusum} methods, to show the modularity of this step.
This final step is further discussed in \Cref{sec:change_detection_methods}.

>>>>

% !TEX root = ../../main.tex
\section{Algorithm overview}\label{sec:method_overview}
The proposed method of this thesis follows the unifying framework as introduced by Takeuchi and Yamanishi \cite{takeuchi2006unifying} and an similar implementation by Camci \cite{camci2010change} with \glspl{svm}.
The unifying framework relates the detection of outliers with change points and divides the whole process in two stages.
The first stage determines the outliers in a time series by giving a score based on the deviation from a learned model, and thereby creates a new time series.
The second stage runs on that new created time series and calculates an average over a window of the outlier scores.
The problem of change detection is then reduced to outlier detection over that average-scored time series.
This method is named \gls{changeFinder} by the authors.
The implementation by Camci, which uses \glspl{svm} to detect changes is named \acrlong{svcpd}.

Whereas \gls{changeFinder} uses a two-stage probability based algorithm, our approach follows \gls{svcpd} by constructing an \gls{svm} over a sliding window.
The \gls{svcpd} algorithm uses the location of new data points in the feature space $\mathcal{F}$ with respect to the hypersphere and the hypersphere's radius $R$ to determine whether the new data point represents a change point.
Our approach is a slight simplification of the algorithm: we only use the updated model properties (\ie the radius $R$) to detect a change.
The \gls{svcpd} also tests every new data point with the current modeled hypersphere, to indicate whether it is an outlier.
This difference is further discussed and justified in \Cref{sec:method_change_detection}.

This section gives a description of the method used for the experiments and change detection mechanism.
First described is the method to process the gathered sensor data.
A schematic overview is given in \Cref{fig:method_overview} and shows the steps of the method.
A more detailed explanation of the ``Update model'' step is given in the remainder of this section.

As graphically represented in \Cref{fig:method_overview}, the change detection method starts by processing the data from sensor, such as the accelerometer, magnetic orientation, and rotation metrics.

The first step is to process the raw streams of data originating from a multiple of sensors.
The two processes applied are alignment and normalization.
Due to noisy sampling, not all the timestamps in the data streams are sensed at the same timestamp.
Since the \gls{svdd} method requires all the data stream at every timestamp and can not handle missing data on one of the timestamps, all the unique timestamps are filtered out.
Whilst this results in an overall filtering effect, in practice between $1\%$ and $5\%$ of each data stream is discarded.
The effect of this filtering is not significant and the data is not modified.

Due to the nature of the sensor signals, a normalization step is required in order to set the weight for all the data streams equal.
The range of the accelerometer signal typically spans $-20$ to $20~m/s^2$, the magnetic field from $-60$ to $60~\mu T$ and the rotations value range is from $-1$ to $1$.
This means that a relative small change in the accelerometer stream could have a much larger impact on the model than the same (absolute) change in the rotation stream, whilst the latter has a larger relative impact.
The normalization step ensures that all data is weighted equally and changes in the data are all proportional.

@@

This final step of this method, step $5$, is the interpretation of the model properties.
Many algorithms can be used for this process, all which take a one-dimensional time series as input and determine where change has occurred.
In our setup we used the \gls{rt} and \gls{cusum} methods, to show the modularity of this step.
This final step is further discussed in \Cref{sec:method_change_detection}.
